Hierarchical recurrent models reveal latent states of neural activity in \textit{C. elegans}

### 2018-01-02 drafts/elife/draft.tex

% Hierarchical model
The third component of our framework is a \emph{hierarchical} model
for learning canonical dynamics from partial recordings collected from
many worms.  Since each recording contains a different subset of
labeled neurons, our first step is to align the data, treating the
activity of neurons that were not found in a given recording as
missing data that must be inferred.  We handle this by integrating
over possible activity of the missing neurons, analytically
marginalizing out this missing data.  The second challenge is
combining information from many recordings while also allowing for
worm-to-worm differences.  For example, the neural activitiy may have
larger variance in some recordings than in others due to different
levels of expression of the fluorescent protein.  We allow for this by
having a separate variance for each worm and neuron, as shown in
Fig.~\ref{fig:model}g.  More importantly, the dynamics of neural
activity may be slightly different from worm to worm.  We account for
this by introducing a canonical library of dynamical regimes
(Fig.~\ref{fig:model}f) that are shared by all worms as well as a
unique library for each individual worm (Fig.~\ref{fig:model}d);
however, we constrain the per-worm dynamics to be close to the
canonical dynamics.

% Inference
We fit the the canonical and per-worm dynamics libraries, the
parameters of the dynamics and observation noise models, and the
discrete and continuous latent states with approximate Bayesian
inference as described in the Supplementary Material. Briefly, we
separate the problem into two steps: learning the continuous latent
states and the mapping from continuous latent states to observations,
then learning the dynamics libraries that govern the continuous latent
states and the correspdonding discrete state sequences.  Both steps are
performed with a block Gibbs sampler, leveraging the structure of the
probabilistic model to perform efficient conditional updates of the
various parameters.

@@

The predicted activity is based on the correlations between neurons
learned from many recordings.  These predictions are a useful aid for
identifying unlabeled neurons.  For example, we see that
\textsf{SMBDL} was not found in the first worm
(Fig.~\ref{fig:emissions}a). If the activity of one of the unlabeled
neurons closely matches the model's predictions for \textsf{SMBDL},
then this provides some evidence in favor of assigning it this
label. We tested this idea by witholding the labels of some
confidently-identified neurons and comparing their activity to the
model's predicted activity for all unlabeled neurons.  We measured the
similarity between the actual and predicted activity and found that
correct label, or ``candidate,'' tends to be much more similar than
other incorrect candidates, as summarized in
Fig.~\ref{fig:emissions}b.  Combined with other side-information, like
the location of the neuron, this provides a powerful aid for identifying
neurons.  The more neurons that we can identify, the better we will be
able to learn about the dynamics of neural activity.

@@

The linear mapping from latent states to observations reveals clusters
of similarly tuned neurons.  The mapping maybe viewed as either an $N$
(neurons) by $D$ (latent dimensions) matrix or as a collection of~$N$
vectors in $D$-dimensional space, i.e. as an embedding of neurons.
These two views are shown in Fig.~\ref{fig:emissions}d. We note that
many vectors point in similar directions, which implies that there are
groups of correlated neurons.  We clustered the neurons based on the
cosine similarity between their embedding vecors and found five clusters
of neurons shown in Fig.~\ref{fig:emissions}f (order of neurons is the same
as in Fig.~\ref{fig:emissions}a). To make sense of these clusters, we then
looked at their spatial distribution along the head-to-tail axis of the
worm's body.  We found that clusters 2-4 correspond to activity closer to
the head and working backward toward the tail.  The last and largest cluster
contains many neurons in the nerve ring (location 0.15) \todo{check} as well
as some posterior neurons. \todo[inline]{discuss functional significance of clusters.}

@@

\begin{figure}[t!]
\centering
\includegraphics[width=6in]{figures/figure3} 
\caption{ \textit{The hierarchical SLDS decomposes continuous states
    into segments with simple linear dynamics.}
  \textbf{a.}~Three dimensions of the inferred continuous latent states
  for the first worm, color coded by the corresponding discrete latent
  state.
  \textbf{b.}~We select the number of discrete states and justify the
  recurrent and robust extensions by measuring the marginal likelihood
  of held-out test data. We find that the robust, recurrent model
  with eight states performs best.
  \textbf{c.}~The inferred segmentations align with the manual segmentations
  of~\citet{kato2015global}.
  \textbf{d.}~Moreover, the inferred states are in close correspodence with
  the manually labeled states, which were shown to indicate different behaviors.
  \textbf{e.}~Each discrete state is associated with a canonical set of linear
  dynamics, which can be viewed as mappings from~$\reals^10$ to~$\reals^10$.
  We show the mappings for the first three dimensions here, with vectors placed
  at locations where each state was used.  We see a variety of rotational
  states that can be combined to yield dynamics like those in~\textbf{a.}.
  \textbf{f.}~We simulate these linear dynamics from a variety of starting
  locations for the fourth (green) and highlight four for closer inspection.
  \textbf{g.}~These four trajectories give rise to different activations of
  the neural clusters. We see see a dip then rise in cluster 3, which consists
  of neurons related to ventral turns, consistent with the manual labeling
  of the green state in~\textbf{d}.
}
\label{fig:syllables}
\end{figure}

>>>>

% Hierarchical model
The third component of our framework is a \emph{hierarchical} model
for learning canonical dynamics from partial recordings collected from
many worms.  Since each recording contains a different subset of
labeled neurons, our first step is to align the data, treating the
activity of neurons that were not found in a given recording as
missing data that must be inferred.  We handle this by integrating
over possible activity of the missing neurons, analytically
marginalizing out this missing data.  The second challenge is
combining information from many recordings while also allowing for
worm-to-worm differences.  For example, the neural activity may have
larger variance in some recordings than in others due to different
levels of expression of the fluorescent protein.  We allow for this by
having a separate variance for each worm and neuron, as shown in
Fig.~\ref{fig:model}g.  More importantly, the dynamics of neural
activity may be slightly different from worm to worm.  We account for
this by introducing a canonical library of dynamical regimes
(Fig.~\ref{fig:model}f) that are shared by all worms as well as a
unique library for each individual worm (Fig.~\ref{fig:model}d);
however, we constrain the per-worm dynamics to be close to the
canonical dynamics.

% Inference
We fit the the canonical and per-worm dynamics libraries, the
parameters of the dynamics and observation noise models, and the
discrete and continuous latent states with approximate Bayesian
inference as described in the Supplementary Material. Briefly, we
separate the problem into two steps: learning the continuous latent
states and the mapping from continuous latent states to observations,
then learning the dynamics libraries that govern the continuous latent
states and the corresponding discrete state sequences.  Both steps are
performed with a block Gibbs sampler, leveraging the structure of the
probabilistic model to perform efficient conditional updates of the
various parameters.

@@

The predicted activity is based on the correlations between neurons
learned from many recordings.  These predictions are a useful aid for
identifying unlabeled neurons.  For example, we see that
\textsf{SMBDL} was not found in the first worm
(Fig.~\ref{fig:emissions}a). If the activity of one of the unlabeled
neurons closely matches the model's predictions for \textsf{SMBDL},
then this provides some evidence in favor of assigning it this
label. We tested this idea by withholding the labels of some
confidently-identified neurons and comparing their activity to the
model's predicted activity for all unlabeled neurons.  We measured the
similarity between the actual and predicted activity and found that
correct label, or ``candidate,'' tends to be much more similar than
other incorrect candidates, as summarized in
Fig.~\ref{fig:emissions}b.  Combined with other side-information, like
the location of the neuron, this provides a powerful aid for identifying
neurons.  The more neurons that we can identify, the better we will be
able to learn about the dynamics of neural activity.

@@

The linear mapping from latent states to observations reveals clusters
of similarly tuned neurons.  The mapping maybe viewed as either an $N$
(neurons) by $D$ (latent dimensions) matrix or as a collection of~$N$
vectors in $D$-dimensional space, i.e. as an embedding of neurons.
These two views are shown in Fig.~\ref{fig:emissions}d. We note that
many vectors point in similar directions, which implies that there are
groups of correlated neurons.  We clustered the neurons based on the
cosine similarity between their embedding vectors and found five clusters
of neurons shown in Fig.~\ref{fig:emissions}f (order of neurons is the same
as in Fig.~\ref{fig:emissions}a). To make sense of these clusters, we then
looked at their spatial distribution along the head-to-tail axis of the
worm's body.  We found that clusters 2-4 correspond to activity closer to
the head and working backward toward the tail.  The last and largest cluster
contains many neurons in the nerve ring (location 0.15) \todo{check} as well
as some posterior neurons. \todo[inline]{discuss functional significance of clusters.}

@@

\begin{figure}[t!]
\centering
\includegraphics[width=6in]{figures/figure3} 
\caption{ \textit{The hierarchical SLDS decomposes continuous states
    into segments with simple linear dynamics.}
  \textbf{a.}~Three dimensions of the inferred continuous latent states
  for the first worm, color coded by the corresponding discrete latent
  state.
  \textbf{b.}~We select the number of discrete states and justify the
  recurrent and robust extensions by measuring the marginal likelihood
  of held-out test data. We find that the robust, recurrent model
  with eight states performs best.
  \textbf{c.}~The inferred segmentations align with the manual segmentations
  of~\citet{kato2015global}.
  \textbf{d.}~Moreover, the inferred states are in close correspondence with
  the manually labeled states, which were shown to indicate different behaviors.
  \textbf{e.}~Each discrete state is associated with a canonical set of linear
  dynamics, which can be viewed as mappings from~$\reals^10$ to~$\reals^10$.
  We show the mappings for the first three dimensions here, with vectors placed
  at locations where each state was used.  We see a variety of rotational
  states that can be combined to yield dynamics like those in~\textbf{a.}.
  \textbf{f.}~We simulate these linear dynamics from a variety of starting
  locations for the fourth (green) and highlight four for closer inspection.
  \textbf{g.}~These four trajectories give rise to different activations of
  the neural clusters. We see see a dip then rise in cluster 3, which consists
  of neurons related to ventral turns, consistent with the manual labeling
  of the green state in~\textbf{d}.
}
\label{fig:syllables}
\end{figure}
